---
title: "Prosper Loans Exploration"
author: "Joseph Thoma"
date: "11/25/2019"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(knitr) 
library(ggthemes) 
library(gridExtra) 
library(dplyr)
library(tidyr) 
library(scales)


```

```{r echo=FALSE, Load_the_Data}
# Load the Data

setwd("D:/Desktop/WGU Projects/data_analyst_nanodegree/data_Analysis_with_R/Final Project")
pl <- read.csv('prosperLoanData.csv')

```

> **Introduction**: This data set contains 113,937 loans with 81 variables on each loan, including loan amount, borrower rate (or interest rate), current loan status, borrower income, borrower employment status, borrower credit history, and the latest payment information.

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
  There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}
str(pl)

summary(pl)



```

Going to start by exploring the Borrower data.

```{r echo=FALSE}
# 1.  Borrowers by State

ggplot(data = pl, aes(x = BorrowerState)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), size = 3, vjust = -1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Which state has the most borrowers?  California exceeds the rest of the states with 14,717 loans.

```{r echo=FALSE}

# 2.  Emplotment Status by Borrower

ggplot(data = pl, aes(x = EmploymentStatus)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), size = 3, vjust = -1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What is the Employment Status of Borrowers?  We see that most borrowers are employed. 
(I did notice that there is 'Employed' then there is 'Full-Time' and 'Part-Time'.)

```{r echo=FALSE}

# 3.  Homeowner per Borrower

ggplot(data = pl, aes(x = IsBorrowerHomeowner)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), size = 3, vjust = -1)
```

Are most borrowers homeowners?  This is pretty even except there are more borrowers that are homeowners.

```{r echo=FALSE}

# 4. Income Range of Borrower

ggplot(data = pl, aes(x = IncomeRange)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), size = 3, vjust = -1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What is the most common income range amoungst borrowers?  We see a normal distribution here with most incomes are within the range of '$25,000 - $74,999'

```{r echo=FALSE}

# 5. Propser Rating of Borrower

ggplot(data = pl, aes(x = ProsperRating..Alpha.)) +
  geom_bar(stat = 'count') +
  geom_text(stat = 'count', aes(label = ..count..), size = 3, vjust = -1)
```

What is the most common Prosper Alpha rating amoungst borrowers?  We see that most borrowers did not have a rating. 
However, we see'C' is the most popular rating for those who have been rated

```{r echo=FALSE}

# 6. Debt to Income Ratio of Borrower

ggplot(aes( x = DebtToIncomeRatio ), data = subset(pl, !is.na(DebtToIncomeRatio))) +
  geom_histogram(binwidth = .25)  + 
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0,10,1))

```

We see that most of the borrowers have a low debt to income ratio 


Next I am going to look at Loan Data
```{r echo=FALSE}

# 1. Loan Amount

ggplot(data = pl, aes(x = LoanOriginalAmount)) +
  geom_histogram ()
  stat_bin(bins = 1000)
```

What is the most popular loan amount?  We see that most loans are under $15,000.
With the majority of loans are $4000.

```{r echo=FALSE}

#2. Loan terms

ggplot(data = pl, aes(x = Term)) +
  geom_bar() +
  stat_bin(bins = 10) +
  geom_text(stat = 'count', aes(label = ..count..), size = 3, vjust = -1)

```

What are the terms of the loans?  Most loans are 36 months.

```{r echo=FALSE}

# 3. Loan Status

ggplot(data = pl, aes(x = LoanStatus)) +
  geom_bar(stat = 'count') +
  geom_text(stat = 'count', aes(label = ..count..), size = 3, vjust = -1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What is the count of each loan status?  Most loans are in the 'current' status.


```{r echo=FALSE}

# 4. Loans per year
pl$LoanOriginationYear <- substring(pl$LoanOriginationDate, 0, 4)

ggplot(aes(x = LoanOriginationYear), data = pl) +
  geom_bar() + 
  geom_text( stat = 'count', aes(label = ..count..), size = 3, vjust = -1)
  
    
```

How many loans issued per year?  We see that 2013 had the most loans.
Noticed a large dip in 2009.  In 2009 Prosperloans were suspended for lending activities pending upon SEC approval. 

```{r echo=FALSE}

# 5. Lender Yield
#https://www.investopedia.com/ask/answers/09/difference-between-yields-and-interest-rate.asp

ggplot(aes( x = LenderYield ), data = pl) +
  geom_bar() +
  scale_y_log10() 
  
    
```

What is the lender yield of loans?


# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?
This dataset consists of 81 variables, with 113937 observations. Dataset is about Prosper Loans from 2005 through 2014.

### What is/are the main feature(s) of interest in your dataset?
I broke down this dataset in two main parts.  Borrowers and Loans.

Borrowers - Most borrowers are from California and made betweend $25k and 75k.  Most borrowers are employed.  It was almost and even split whether borrowers owned their own home or not.  Most borrowers had a good debt to Income ratio.

Loans - Most loans were under $15,00o and at 36 month term. Most loans were in good standing.  Between 2005 and 2014, 2013 had the most loans issued.  There was also a bid decline in loans in 2009 because Prosper Loans lending activities were suspended.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
I will need to look into Borrower Rate and Borrower APR.  

### Did you create any new variables from existing variables in the dataset?
I created 'LoanOriginationYear' out of 'LoanOriginationDate'.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
I noticed that withing the 'Employment status' that there were the following categories, 'Employed', 'Full Time', and 'Part Time'.  These three seem to be the same and probably can be combined   .

I also noticed that in the Prosper Rating that there were a lot of nulls.


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

# 1. 
```



```{r echo=FALSE, Bivariate_Plots}

# 2. 


```


```{r echo=FALSE, Bivariate_Plots}

# 3. 

```


```{r echo=FALSE, Bivariate_Plots}

# 4. 

```


```{r echo=FALSE, Bivariate_Plots}

# 5. 

```
# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
  
  ### Did you observe any interesting relationships between the other features \
  (not the main feature(s) of interest)?
  
  ### What was the strongest relationship you found?
  
  
  # Multivariate Plots Section
  
  > **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
                                                                         units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------
  
  # Reflection
  
  > **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!